---
title: "Supplementary - Data Cleaning and Visualization"
author: "Meghan Myles and Owen Fiore"
date: "2023-12-17"
output: html_document
---
  
  # Data Setup and Cleaning
  
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(foreign)
```

```{r}
df <- read.csv("RawData.csv")
```

## Data Cleaning

```{r}
original_nrow <- nrow(df)
print(paste('Original number of rows:', original_nrow))

# Convert blank cells to NA
df[df == ""] <- NA
df[df == " "] <- NA
df <- na.omit(df)
na_omitted_nrow <- nrow(df)
print(paste('Number of rows after dropping rows with null values:', na_omitted_nrow))
print(paste('Number of rows dropped:', original_nrow - na_omitted_nrow))
```

We dropped 11,398 rows, leaving us with 19,243 rows. This is acceptable. We will now check for duplicate rows.

```{r}
duplicate_rows <- df[duplicated(df), ]
print("Duplicate Rows:")
print(duplicate_rows)

cleaned_df <- df # store a version of the df that is not one-hot encoded
cleaned_df$AggregateScore <- cleaned_df$MathScore + cleaned_df$ReadingScore + cleaned_df$WritingScore
```

There are no duplicate rows.

We will now explore some of our categorical variables, and change strings to numerical labels. We will start with our binary variables.

We will encode female as '0' and male as '1'. We will also change the name of the 'Gender' column to 'Male' to better reflect this change.

```{r}
Gender_counts <- df %>%
  count(Gender)
print(Gender_counts)

df <- df %>%
  mutate(Gender = ifelse(Gender == 'male', 1, 0))

names(df)[names(df) == "Gender"] <- "Male"
```

We will encode LunchType: 'free/reduced' as 1 and LunchType: 'standard' as 0. We will also change the name of this column to 'LunchDiscount' to better reflect the changes made.

```{r}
LunchType_counts <- df %>%
  count(LunchType)
print(LunchType_counts)

df <- df %>%
  mutate(LunchType = ifelse(LunchType == 'free/reduced', 1, 0))
names(df)[names(df) == "LunchType"] <- "LunchDiscount"
```

We will encode TestPrep: completed as 1 and TestPrep: none as 0.

```{r}
TestPrep_counts <- df %>%
  count(TestPrep)
print(TestPrep_counts)

df <- df %>%
  mutate(TestPrep = ifelse(TestPrep == 'completed', 1, 0))
```

We will encode IsFirstChild: yes as 1, and IsFirstChild: no as 0.

```{r}
IsFirstChild_counts <- df %>%
  count(IsFirstChild)
print(IsFirstChild_counts)

df <- df %>%
  mutate(IsFirstChild = ifelse(IsFirstChild == 'yes', 1, 0))
```

We will encode TransportMeans: private as 1, and TransportMeans: school_bus as 0. We will also change the column name to 'PrivateTransport' to better reflect these changes.

```{r}
TransportMeans_counts <- df %>%
  count(TransportMeans)
print(TransportMeans_counts)

df <- df %>%
  mutate(TransportMeans = ifelse(TransportMeans == 'private', 1, 0))
names(df)[names(df) == "TransportMeans"] <- "PrivateTransport"
```

As a review, we have made the following encoding changes:
  
  | Original variable name | New variable name | 1         | 0      |
  |------------------------|-------------------|-----------|--------|
  | Gender                 | Male              | male      | female |
  | LunchType              | LunchDiscount     | yes       | no     |
  | TestPrep               |                   | completed | none   |
  | IsFirstChild           |                   | yes       | no     |
  | TransportMeans         | PrivateTransport  | yes       | no     |
  
  Now, we will use one-hot encoding to encode the non-binary categorical variables.

```{r}
EthnicGroup_counts <- df %>%
  count(EthnicGroup)
print(EthnicGroup_counts)

df <- df %>%
  mutate(
    EthnicGroupA = ifelse(EthnicGroup == 'group A', 1, 0),
    EthnicGroupB = ifelse(EthnicGroup == 'group B', 1, 0),
    EthnicGroupC = ifelse(EthnicGroup == 'group C', 1, 0),
    EthnicGroupD = ifelse(EthnicGroup == 'group D', 1, 0),
    EthnicGroupE = ifelse(EthnicGroup == 'group E', 1, 0)
  )

df <- df %>%
  select(-EthnicGroup)

head(df)
```
Now we convert ParentEduc to be a factor variable as its levels are ordinal
```{r}
ParentEduc_counts <- df %>%
  count(ParentEduc)
print(ParentEduc_counts)

df$ParentEduc <- as.factor(df$ParentEduc)
df$ParentEduc <- factor(df$ParentEduc, levels = c(
  "some high school", "high school", "some college",
  "associate's degree", "bachelor's degree", "master's degree"
))

df$ParentEduc <- as.integer(df$ParentEduc)

head(df)
```

Now we one hot encode ParentMaritalStatus
```{r}
ParentMaritalStatus_counts <- df %>%
  count(ParentMaritalStatus)
print(ParentMaritalStatus_counts)

df <- df %>%
  mutate(
    ParentDivorced = ifelse(ParentMaritalStatus == "divorced", 1, 0),
    ParentMarried = ifelse(ParentMaritalStatus == "married", 1, 0),
    ParentSingle = ifelse(ParentMaritalStatus == "single", 1, 0),
    ParentWidowed = ifelse(ParentMaritalStatus == "widowed", 1, 0)
  )

df <- df %>%
  select(-ParentMaritalStatus)

head(df)
```

We convert PracticeSport to a factor variables as its levels are ordinal
```{r}
PracticeSport_counts <- df %>%
  count(PracticeSport)
print(PracticeSport_counts)

df$PracticeSport <- factor(df$PracticeSport, levels = c("never", "sometimes", "regularly"))
df$PracticeSport <- as.integer(df$PracticeSport)

head(df)
```

We convert WklyStudyHours to a factor variables as its levels are ordinal
```{r}
WklyStudyHours_counts <- df %>%
  count(WklyStudyHours)
print(WklyStudyHours_counts)


df$WklyStudyHours <- factor(df$WklyStudyHours, levels = c("5 - 10", "< 5", "> 10"))
df$WklyStudyHours <- as.integer(df$WklyStudyHours)

head(df)
```

Now, all of our variables have been one-hot encoded. We will move the response variable columns to the end of the df for readability.

```{r}
df <- df %>%
  select(-MathScore, -ReadingScore, -WritingScore,
         MathScore, ReadingScore, WritingScore)

head(df)
```

As a reminder, we have made the following changes:
  
  | Original Column name | New Created Columns      |
  |----------------------|--------------------------|
  | EthnicGroup          | EthnicGroupA             |
  |                      | EthnicGroupB             |
  |                      | EthnicGroupC             |
  |                      | EthnicGroupD             |
  |                      | EthnicGroupE             |
  | ParentMaritalStatus  | ParentDivorced           |
  |                      | ParentMarried            |
  |                      | ParentSingle             |
  |                      | ParentWidowed            |
  
  We will create a new calculated field 'AggregateScore' representing each student's aggregate Math, Reading, and Writing scores.
```{r}
df['AggregateScore'] <- df['MathScore'] + df['ReadingScore'] + df['WritingScore']

head(df)
```


## Data Visualization

We will create some basic visualizations of our data in order to better understand its structure.
```{r}
par(mfrow = c(2, 2))
hist(df$MathScore, main = "Math Score Distribution", xlab = "Math Score")
hist(df$ReadingScore, main = "Reading Score Distribution", xlab = "Reading Score")
hist(df$WritingScore, main = "Writing Score Distribution", xlab = "Writing Score")
hist(df$AggregateScore, main = "Aggregate Score Distribution", xlab = "Aggregate Score")
```


```{r}
print(unique(cleaned_df$ParentEduc))
```

We can see that each of the above histograms is unimodal and left-skew.

```{r}
par(pin = c(100, 100))
par(mfrow = c(3, 2), mar = c(2, 4, 2, 2))

boxplot(AggregateScore ~ Gender, data = cleaned_df,
        main = "Aggregate Scores by Gender",
        xlab = "Gender",
        ylab = "Scores",
        col = c("pink", "skyblue"))

boxplot(AggregateScore ~ LunchType, data = cleaned_df,
        main = "Aggregate Scores by Lunch Type",
        xlab = "Lunch Type",
        ylab = "Scores",
        col = c("#A2C5AC", "#7F6A93"))

boxplot(AggregateScore ~ TestPrep, data = cleaned_df,
        main = "Aggregate Scores by Test Preparation Status",
        xlab = "Test Preparation Status",
        ylab = "Scores",
        col = c("#00537a", "#ffba42"))

order_of_plots <- c("never", "sometimes", "regularly")
cleaned_df$PracticeSport <- factor(cleaned_df$PracticeSport, levels = order_of_plots)
boxplot(AggregateScore ~ PracticeSport, data = cleaned_df,
        main = "Aggregate Scores by Sports Practice",
        xlab = "Sports Practice",
        ylab = "Scores",
        col = c("#fd0363", "#9c1057", "#3b1d4a"))

boxplot(AggregateScore ~ IsFirstChild, data = cleaned_df,
        main = "Aggregate Scores by First Child Status",
        xlab = "First child?",
        ylab = "Scores",
        col = c("#afdedc", "#fdd54f"))

boxplot(AggregateScore ~ TransportMeans, data = cleaned_df,
        main = "Aggregate Scores by School Transportation",
        xlab = "Uses private transport?",
        ylab = "Scores",
        col = c("#ffadad", "#bdb2ff"))

```

```{r}
par(pin = c(100, 100))
par(mfrow = c(3, 2), mar = c(2, 4, 2, 2))

boxplot(AggregateScore ~ NrSiblings, data = cleaned_df,
        main = "Aggregate Scores by Number of Siblings",
        xlab = "Number of siblings",
        ylab = "Scores",
        col = c("#f55c7a", "#f56c77", "#f57c73", "#f68c70", "#f69c6d", "#f6ac69", "#f6bc66", "#EED329"))


order_of_plots <- c("some high school", "high school", "some college", "associate's degree", "bachelor's degree", "master's degree")
cleaned_df$ParentEduc <- factor(cleaned_df$ParentEduc, levels = order_of_plots)
boxplot(AggregateScore ~ ParentEduc, data = cleaned_df,
        main = "Aggregate Scores by Parent Education Level",
        xlab = "Parental Education Level",
        ylab = "Scores",
        col = c("#422680", "#341671", "#280659", "#660f56", "#ae2d68", "#f54952"))

boxplot(AggregateScore ~ WklyStudyHours, data = cleaned_df,
        main = "Aggregate Scores by Weekly Study Hours",
        xlab = "Weekly Study Hours",
        ylab = "Scores",
        col = c("#3d0066", "#c86bfa", "#fdc500"))

boxplot(AggregateScore ~ ParentMaritalStatus, data = cleaned_df,
        main = "Aggregate Scores by Parent Marital Status",
        xlab = "Parental Marital Status",
        ylab = "Scores",
        col = c("#219ebc", "#023047", "#ffb703", "#fb8500"))

boxplot(AggregateScore ~ EthnicGroup, data = cleaned_df,
        main = "Aggregate Scores by Ethnic Group",
        xlab = "Ethnic Group",
        ylab = "Scores",
        col = c("#2c78b7", "#54c5e0", "#a3ebe7", "#f0f6f6", "#ffafd7"))
```

In looking at the above 10 boxplots, we can visually compare aggregate score means between groups. This will give us an indication of what variables we expect to vary significantly between groups. It would appear that variables which correlate with different scores across groups include gender, whether or not a student received a lunch discount, whether or not a student completed test preparation, highest parental education level, and ethnic group. The variables which do not seem to have variable average test scores across groups are whether or not a student participates in sports, whether or not a student is a first child, number of siblings, weekly study hours, and parental marital status.
